---
title: "leca_circleplot"
author: "Rachael M. Cox"
date: "3/22/2022"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(network)
library(igraph)
library(tidygraph)
library(ggraph)
theme_set(theme_bw(base_size = 12))
pal_pretty <- c("#0072B2","#E69F00","#009E24","#FF0000", "#5530AA")
pal_npg <- c("#E64B35", "#4DBBD5", "#00A087", "#3C5488",
                 "#F39B7F", "#8491B4", "#91D1C2", "#DC0000",
                 "#7E6148", "#B09C85")
## the custom function using Color Brewer
cols_f <- colorRampPalette(RColorBrewer::brewer.pal(8, 'Spectral'))
spectral <- as.vector(cols_f(8))

workdir <- "/stor/work/Marcotte/project/rmcox/leca/"
```

## Load data

//**TODO**: Cannot get semicolon-delimited entries formatted correctly; example: "H4C1; H4C2; H4C3; H4C4; H4C5;" ...

```{r}
# clusters & complexes
clst_annot_file <- paste0(workdir, "ppi_ml/results/ppi_clustering/final_cmplx_annots_011823.csv") 

clst_annot <- read_csv(clst_annot_file) %>%
  mutate(human_family_size = str_count(human_gene_names_primary, ',')+1) %>%
  mutate(human_family_size = ifelse(is.na(human_family_size), 0, human_family_size)) %>% 
  #mutate(human_genes_fmt = gsub(";", ",", human_gene_names_primary, fixed=TRUE)) %>%
  mutate(human_genes_fmt = str_replace_all(human_gene_names_primary, pattern = ";", replacement = ",")) %>% 
  mutate(human_genes_fmt = str_replace(human_gene_names_primary, "^([^,]*,[^,]*),.*", "\\1 ...")) %>%
  mutate(human_genes_fmt = str_replace(human_genes_fmt, ", ", "\n")) %>%
  mutate(genes_fmt = ifelse(human_family_size != 0 & human_family_size <= 5, human_genes_fmt, ID)) %>% 
  select(ID, human_gene_names_primary, human_genes_fmt, human_family_size, genes_fmt,
         matches("cut_*"), notes, granulated_cmplx_name, characterization_status)
```

## Data exploration

#### Influence of big orthogroups
A significant portion of the conserved interactome is based on orthogrouped peptide measurements. Some of these orthogroup are massive families, potentially skewing interpretation of the interaction results. The data exploration below attempts to determine if and how we should flag it.

```{r}
# summary stats
summary(clst_annot$human_family_size)

# manual counts
levels = c("0", "1", "2", "3", "4", "5", ">=6")
clst_annot_exp <- clst_annot %>%
  mutate(family_size_binned = ifelse(human_family_size >= 6, ">=6", as.character(human_family_size)))
# lock in var order
clst_annot_exp$family_size_binned <- factor(clst_annot_exp$family_size_binned, levels = levels)



# distribution of human family sizes per orthogroup
ggplot(clst_annot_exp, aes(x = human_family_size)) +
  geom_histogram(binwidth = 1) +
  labs(x = "Human proteins per orthogroup")

# binned counts (most useful I think)
ggplot(clst_annot_exp, aes(x = family_size_binned)) +
  geom_bar() +
  geom_text(stat = "count", aes(label=..count..), vjust = -0.25) +
  labs(x = "Human proteins per orthogroup")
```

Approximately 10% (257/2778) of high-scoring LECA orthogroups represent the measurement of large families (>=6 proteins, in humans at least). **Might be worth generating the # of proteins per orthogroup for all involved species and getting an average.**

#### Complex annotation status

Manual labeling of PPI characterization status is in the works; the code below indicates this progress:

```{r}
clst_char <- clst_annot %>%
  group_by(characterization_status) %>% 
  tally() %>%
  arrange(desc(n))  # 01/10/23: still 1373 NAs out of 2778; 01/18/23: 61/2778 NAs (97.8% labeled)

clst_cmplx <- clst_annot %>%
  group_by(granulated_cmplx_name) %>% 
  tally() %>%
  arrange(desc(n))
```

## Prep data for circle plot

```{r}
# filter data
pdf <- clst_annot %>%
  filter(!is.na(characterization_status)) %>% 
  mutate(characterization_status_fmt = case_when(characterization_status == "Known" ~ "Known protein complex",
                                             characterization_status == "Novel member" ~ "Novel association",
                                             characterization_status == "Uncharacterized" ~ "Uncharacterized protein complex",
         is.na(characterization_status) ~ "Unclear")) 
```

Functions:
```{r}
format_clst <- function(clusters, level_vect){
  # get clusters greater than one big
  highest_level <- level_vect[1]
  clusters_filt <- clusters %>%
    add_count(!!sym(highest_level)) %>%
    filter(n > 1) %>%
    select(-n)
  # make each cluster number unique by adding on the header
  clusters_uniqued_long <- clusters_filt %>%
    gather(clusterset, clusternum, -ID) %>%
    mutate(clusterid = paste0(clusternum, clusterset)) %>% 
    select(-clusternum)
  return(clusters_uniqued_long)
}

make_edge <- function(n, clu, cols){
    startcol = cols[n]
    endcol = cols[n + 1]
    clu %>% select(!!startcol, !!endcol) %>% rename(from = !!startcol, to = !!endcol)
}

graph_clst <- function(clusters_uniqued_long, level_vect) {
  # make wide
  clusters_uniqued <- clusters_uniqued_long %>%
    spread(clusterset, clusterid)
  # set origin  
  clusters_uniqued$link <- "origin"
  # get origin col, cut cols, and ID col
  clusters_uniqued_sel <- clusters_uniqued %>%
    select(link, {{ level_vect }}, ID)
  # get col names
  cols <- names(clusters_uniqued_sel)
  # generate edges
  edges <-
    map_df(seq(1, length(cols) - 1), make_edge, clusters_uniqued_sel, cols) %>%
    unique()
  # generate graph
  mygraph2 <- as_tbl_graph(edges)
  return(mygraph2)
}

#no bordering
# Idea of this is to label only largest sets that label applies to 
# Also set that complex name as the label value for one member of largest set that that label applies to
format_labels <- function(clusters, annot, level_vect){
  
  # First get only selected clustering levels
  clusterid_annot <- clusters %>% left_join(annot, by = "ID") %>%
    filter(clusterset %in% {{ level_vect }}) %>%
    select(clusterset, clusterid, complex_label) %>%
  
    # Count numbers of members per cluster set
    group_by(clusterid) %>%
      mutate(unique_types = n_distinct(complex_label)) %>%
      add_count(clusterset, complex_label, sort = TRUE) %>%
    
    group_by(complex_label) %>%
      filter(unique_types ==  min(unique_types)) %>%
      mutate(label = case_when(row_number() == 1 ~ complex_label)) %>%
      filter(!is.na(label))  %>%
    group_by(clusterid) %>%
      summarize(label = paste(unique(label), collapse = ', ')) %>% 
    ungroup()

  return(clusterid_annot)
}

annotate_graph <- function(graph, clusterid_annot, annot){
  graph_annot <- graph %>% activate('nodes') %>%
    left_join(clusterid_annot, by = c("name"  = "clusterid")) %>%
    left_join(annot, by = c("name"  = "ID")) 
  return(graph_annot)

}

plot_graph <- function(graph, labels = TRUE){

  circlepack_plants_highlight <- ggraph(graph, layout = 'circlepack') + 
    geom_node_circle(aes(  fill = set), size = 0.1, color = "grey50") +
    scale_fill_manual(values = c(palette[4],  "yellow"), na.value = "white") + #light green "#c5f7d1
    theme(legend.key.size = unit(0.2, "lines"),
          legend.position = "bottom",
          legend.text = element_text(size = 4)) +
    theme_void() + 
    NULL

    if(labels == TRUE){
      # label nodes
      circlepack_plants_highlight <- circlepack_plants_highlight + 
        geom_node_text( aes( label= label), repel = TRUE,  segment.color = "grey40", size = 2, force = 0.5 ) +
        NULL
    }

  circlepack_plants_highlight_nolegend <- circlepack_plants_highlight + 
    theme(legend.position="FALSE")  
  return(circlepack_plants_highlight_nolegend)
}

circlepack_fxn <- function(graph_annot, seed){

 
  set.seed(seed = seed)
  base_layout_circlepack <- ggraph(graph_annot, layout = 'circlepack')

  base_layout_circlepack_format <- base_layout_circlepack +
    geom_node_circle(aes(fill = characterization_status_fmt)) +  # node fill variable
    scale_fill_manual(values = c("#E64B35", "#4DBBD5", "#F0E442", "grey90", "#3C5488", "#00A087"),
                      na.translate=F) + # node fill colors
    scale_color_manual(values = c("grey50") ,na.value = "grey50") +
    scale_size_manual(values = c(1), na.value = 0.5) + # set carefully; big nums freeze RStudio
    theme_void() +
    theme(legend.title = element_blank(),
          legend.position = "top") +
    guides(fill=guide_legend(nrow=1,byrow=TRUE)) #+
    #geom_node_text(aes(label = gene_labels), size = 0.75, color = "black", fontface = "bold")  # node labels
    

  #circlepackplot_clusterlabel <- base_layout_circlepack_format +
      #geom_node_text(aes(label= label), size = 1.5, color = "#400080") + #label is complex label
      #geom_node_text(aes(label= arath_genename), size = 1.5, color = "blue") + #arath_genename is labeling node by gene name
        #theme(legend.position="FALSE")+
        #NULL
  return(base_layout_circlepack_format)
}
```

## Make circle plot

```{r}
# get cut cols
cuts <- names(clst_annot[grepl("cut", names(clst_annot))])
cuts_sel <- cuts[c(3,6,9)]

# format chosen clusters
clusters_uniqued_long <- format_clst(pdf, cuts_sel)

# generate graph objct
graph <- graph_clst(clusters_uniqued_long, cuts_sel)

# join labels
graph <- graph %>% activate("nodes") %>%
  left_join(pdf, by = c("name" = "ID"))

# plot graph object
p <- circlepack_fxn(graph, 6)

# save output
figure_dir <- paste0(workdir, "ppi_ml/figures/")
#ggsave(paste0(figure_dir, "cplot_annot-status_021623.pdf"), device = "pdf", width = 8, height = 8, units = "in")
#ggsave(paste0(figure_dir, "cplot_annot-status_021623.png"), device = "png", width = 8, height = 8, units = "in")

ggsave(paste0(figure_dir, "cplot_annot-status_021623.svg"), device = "svg", width = 8, height = 8, units = "in")

# get legend
legend <- cowplot::get_legend(p)
grid.newpage()
grid.draw(legend)
ggsave(paste0(figure_dir, "cplot_annot-status_legend_021623.svg"), device = "svg", width = 8, height = 8, units = "in")
```

## Test plot params

```{r}
# get cut cols
cuts <- names(clst_annot[grepl("cut", names(clst_annot))])
cuts_sel <- cuts[c(3,6,9)]

# get subset
test <- head(pdf, 500)

# format chosen clusters
clusters_uniqued_long <- format_clst(test, cuts_sel)

# generate graph objct
graph <- graph_clst(clusters_uniqued_long, cuts_sel)

# join labels
graph <- graph %>% activate("nodes") %>%
  left_join(pdf, by = c("name" = "ID"))

# plot graph object
circlepack_fxn(graph, 6)

# save output
figure_dir <- paste0(workdir, "ppi_ml/figures/")
ggsave(paste0(figure_dir, "cplot_param_test.pdf"), device = "pdf", width = 8, height = 8, units = "in")
ggsave(paste0(figure_dir, "cplot_param_test.png"), device = "png", width = 8, height = 8, units = "in")
```

## Color by complex

```{r}

circlepack_fxn <- function(graph_annot, seed){

 
  set.seed(seed = seed)
  base_layout_circlepack <- ggraph(graph_annot, layout = 'circlepack')

  base_layout_circlepack_format <- base_layout_circlepack +
    geom_node_circle(aes(fill = cmplx_label)) +  # node fill variable
    scale_fill_manual(values = c("#E64B35", "#4DBBD5", "#F0E442", "grey90", "#3C5488", "#00A087"),
                      na.translate=F) + # node fill colors
    scale_color_manual(values = c("grey50") ,na.value = "grey50") +
    scale_size_manual(values = c(1), na.value = 0.5) + # set carefully; big nums freeze RStudio
    theme_void() +
    theme(legend.title = element_blank(),
          legend.position = "top") +
    guides(fill=guide_legend(nrow=1,byrow=TRUE))
  
  return(base_layout_circlepack_format)
}

cmplx_labels <- pdf %>%
  pull(granulated_cmplx_name) %>%
  unique()

test <- pdf %>%
  mutate(cmplx_label = ifelse(grepl("\\(ER\\)-localized multiprotein complex", granulated_cmplx_name), granulated_cmplx_name, NA))

# format chosen clusters
clusters_uniqued_long <- format_clst(test, cuts_sel)

# generate graph objct
graph <- graph_clst(clusters_uniqued_long, cuts_sel)

# join labels
graph <- graph %>% activate("nodes") %>%
  left_join(test, by = c("name" = "ID"))

# plot graph object
circlepack_fxn(graph, 6)

figure_dir <- paste0(workdir, "ppi_ml/figures/cplot_cmplx_labels/")
write_delim(cmplx_labels, paste0(figure_dir, "cmplx_list.txt"), delim="\n", col_names = F)

ggsave(paste0(figure_dir, "cplot_label_test.png"), device = "png", width = 8, height = 8, units = "in")


```

## EXAMPLE CODE

```{r}
# # example input file
# momo_example <- read_csv('/stor/work/Marcotte/project/Momo/rbc_figures/circlepack/category_annotated_modified_justrbc.csv') 
# 
# #label each circle with gene names
# momo_label_ex <- read_csv("/stor/work/Marcotte/project/Momo/diffusion_clustering/human_proteome_uiprotID_genename.csv")
# momo_label_ex <- momo_label_ex %>% 
#   group_by(ID) %>% 
#   mutate(num=n()) %>% 
#   ungroup() %>% 
#   filter(num==1)
# 
# # make border annotations
# momo_bordering_ex  <- read_csv('/stor/work/Marcotte/project/Momo/rbc_figures/circlepack/category_annotated_modified_justrbc_bordering_oxidoreductase.csv')
# 
# # pick clst cols
# sel <- names(momo_example)[c(3,5,8)] 
# 
# clusters_uniqued_long <- format_clusters(momo_example, sel)
# 
# graph <- make_graph_clust(clusters_uniqued_long, sel)
# 
# graph <- graph %>% activate("nodes") %>%
#   left_join(momo_label_ex, by = c("name" = "ID"))%>%
#   left_join(momo_bordering_ex, by = c("name"  = "ID")) 
# circlepack_fxn(graph, 68)
# 
# # save output
# ggsave("/stor/work/Marcotte/project/Momo/rbc_figures/circlepack/circlepackplot_category_cut358_rbconly_4cats_cb_friendly.pdf", device = "pdf", width = 6, height = 6, units = "in")
```